Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-11-01 text/ms.tex
Minor edits

Black swans are unexpected extreme events with potentially dramatic
consequences \citep{taleb2007,sornette2009}. One of the most striking black
swans in ecology is the asteroid marking the mass extinctions of the K-T
boundary TODO-REF \citep{harnik2012}. Today, climate extremes, in concert with
shifts in mean temperature, are expected to cause the greatest ecological and
societal damage \citep{ipcc2012}. But, while extremes in the physical
environment such as wave height, storm severity, and temperature are are
frequent experiences \citep{gaines1993,katz2005}, it remains unclear the extent
to which ecological systems buffer or suffer from black swans
\citep{nunez2012}.

@@

We obtained abundance time series from the Global Population Dynamics Database
(GPDD) \citep{gpdd2010}. The GPDD contains nearly 5000 time series of abundance
across $\sim$1000 species and $\sim$100 taxonomic orders. To derive
a high-quality subset of populations that were suitable for our analyses, we
filtered the data (Supporting Material). Our filtering removed populations from
less reliable data sources, removed those without sufficient data for our
models, and interpolated some missing values \citep[\textit{sensu}][]{brook2006a}.
Our final dataset contained \NPops~populations across \NOrders~taxonomic orders
and seven taxonomic classes with a median of \medianTimeSteps~time steps (range
of \minTimeSteps--\maxTimeSteps) (Table S1, Fig.~\ref{fig:all-ts}).

@@

We estimated all models in a Bayesian framework using Stan \citep[][version
2.4.0]{stan-manual2014} via the R computing environment \citep{r2014}. Stan
samples from the posterior distribution with an adaptive version of Hamiltonian
Markov chain Monte Carlo called the No-U-Turn Sampler and generally obtains
less correlated samples than algorithms such as the Gibbs sampler
\citep{hoffman2014}. We assured that chains had sufficiently converged and the
sampler had obtained sufficient independent samples from the posterior
($\widehat{R} < 1.05$, $n_\mathrm{eff} > 200$; Supporting Material). We used
simulated data to test how easily we could detect $\nu$ given different sample
sizes and to ensure we could recover unbiased parameter estimates from the
Gompertz model (Supporting Material).

We chose weakly informative priors to incorporate our understanding of
plausible population dynamics 
(\citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Material). 
For $\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at
values above two (Fig.~\ref{fig:priors}) --- a slightly less informative prior
than suggested by \citet{fernandez1998}. This prior gives only a $7.7$\%
probability that $\nu < 10$ but constrains the sampling sufficiently to avoid
wandering off towards infinity --- above approximately $\nu = 20$ the
t distribution is so similar to the normal distribution that time series of the
length considered here are unlikely to be informative about the precise value
of $\nu$ (Fig.~\ref{fig:didactic}).

We fit alternative population models to test if four key phenomena
systematically changed our conclusions. (1) Autocorrelation has been suggested
as a reason for increased observed variability of abundance time series through
time, which could create apparent heavy tails \citep{inchausti2002}; therefore,
we fit a model that modelled the serial correlation in the residuals. (2)
Previous work has modelled abundance or growth rates without accounting for
density dependence \citep{halley2002,segura2013}; therefore, we fit a simpler
model in which we assumed density independence. (3) Observation error could
bias parameter estimates \citep{knape2012} or mask our ability to detect heavy
tails \citep{ward2007}; therefore, we fit a model where we allowed for a fixed
quantity of observation error ($0.2$ standard deviations on a log scale). (4)
The Gompertz model assumes that population growth rate declines linearly with
log abundance. We also fit an alternative model, the Ricker-logistic model,
which assumes that population growth rate declines linearly with abundance
itself (Supporting Material).

@@

We investigated possible covariates of heavy-tailed population dynamics
visually and through multilevel modelling. We plotted characteristics of the
time series ($\sigma$, $\lambda$, $b$, and time-series length) along with two
life-history characteristics (body length and maximum lifespan obtained from
\citet{brook2006a}) against our estimated probability of heavy tails, Pr$(\nu
< 10) > 0.5$. We formally investigated these relationships by fitting beta
regression multilevel models. For covariates that were derived from Gompertz
model parameter estimates, we incorporated standard deviations around the
means. To account for broad patterns of phylogenetic relatedness, we allowed
for varying intercepts at the taxonomic class, order, and species level
(Supporting Material).

@@

Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy
directions. First, ecological resource management can draw from other
disciplines that focus on heavy tails. For example, earthquake preparedness and
response is focussed on black swan events. Similarly to ecological black swans,
we can rarely predict the specific location and timing of large earthquakes.
But, earthquake preparedness involves spatial planning based on forecast
probabilities to focus early detection efforts and develop disaster response
plans (e.g.~REF-TODO). A similar focus might benefit resource management once
our ability to predict the spatial probability and covariates of ecological
black swans improves. Second, we can develop management policy that is robust
to heavy tails. For example, setting target population abundances that are
appropriately set back from critical limits may buffer black swan events
\citep[e.g.][]{caddy1996}, and maintaining genetic, phenotypic, and behavioural
diversity may allow some components of populations persist when others are
affected by disease or extreme environmental forces
\citep[e.g.][]{schindler2010, anderson2014}. Finally, extreme and unexpected,
surprising, or counterintuitive ecological dynamics offer a tremendous
opportunity to learn about ecological systems, evaluate when our models break
down, and adjust future management policy \citep{doak2008, pine-iii2009,
  lindenmayer2010}. In recent decades ecology has moved toward focussing on
aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a,
  thompson2013}. Our results suggest that an added focus on ecological extremes
represents the next frontier, particularly in the face of increased climate
extremes \citep{meehl2004,ipcc2012}.

@@

\section{Supplementary Material}

The following supplementary material is available online for this article:\\
Figures S1--S11\\
Table S1\\
Example Stan code for a heavy-tailed Gompertz model and multilevel beta
regression\\
The GPDD IDs used in our analysis\\
All code and data to recreate our analysis are available at:\\
\url{https://github.com/seananderson/heavy-tails}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{../analysis/correlates-p10.pdf}
\caption{
Potential covariates of heavy-tailed population dynamics (indicated by a high
probability that $\nu < 10$). Shown are (a--c) parameters from the Gompertz
heavy-tailed population model ($\sigma$, $\lambda$, $b$), (b) number of time
steps, (c) body length, and (d) lifespan. For the Gompertz parameters, $\sigma$
refers to the scale parameter of the Student-t process-noise distribution,
$\lambda$ refers to the expected log abundance at the next time step at an
abundance of one, $b$ refers to the density dependence parameter ($1$ is
maximally density independent, $0$ is maximally density dependent, and $<0$ is
inversely density dependent).
%For parameter values, points and lines represent medians and 50\% credible
%intervals of the posterior.
Circles representing a few sharks, crustaceans, and gastropods are filled in
white. Median and 90\% credible interval posterior predictions of a beta
regression multilevel model are shown in panels a and d where there was a high
probability the slope coefficient was different from zero (Supporting
Material).
}
\label{fig:correlates}
\end{center}
\end{figure}

>>>>

Black swans are unexpected extreme events with potentially dramatic
consequences \citep{taleb2007,sornette2009}. One of the most striking black
swans in ecology is the asteroid marking the end-Cretaceous mass extinction 65 million
years ago \citep{alvarez1980,harnik2012}. Today, climate extremes, in concert with
shifts in mean temperature, are expected to cause the greatest ecological and
societal damage \citep{ipcc2012}. But, while extremes in the physical
environment such as wave height, storm severity, and temperature are are
frequent experiences \citep{gaines1993,katz2005}, it remains unclear the extent
to which ecological systems buffer or suffer from black swans
\citep{nunez2012}.

@@

We obtained abundance time series from the Global Population Dynamics Database
(GPDD) \citep{gpdd2010}. The GPDD contains nearly 5000 time series of abundance
across $\sim$1000 species and $\sim$100 taxonomic orders. To derive
a high-quality subset of populations that were suitable for our analyses, we
filtered the data (Supporting Information). Our filtering removed populations from
less reliable data sources, removed those without sufficient data for our
models, and interpolated some missing values \citep[\textit{sensu}][]{brook2006a}.
Our final dataset contained \NPops~populations across \NOrders~taxonomic orders
and seven taxonomic classes with a median of \medianTimeSteps~time steps (range
of \minTimeSteps--\maxTimeSteps) (Table S1, Fig.~\ref{fig:all-ts}).

@@

We estimated all models in a Bayesian framework using Stan \citep[][version
2.4.0]{stan-manual2014} via the R computing environment \citep{r2014}. Stan
samples from the posterior distribution with an adaptive version of Hamiltonian
Markov chain Monte Carlo called the No-U-Turn Sampler and generally obtains
less correlated samples than algorithms such as the Gibbs sampler
\citep{hoffman2014}. We assured that chains had sufficiently converged and the
sampler had obtained sufficient independent samples from the posterior
($\widehat{R} < 1.05$, $n_\mathrm{eff} > 200$; Supporting Information). We used
simulated data to test how easily we could detect $\nu$ given different sample
sizes and to ensure we could recover unbiased parameter estimates from the
Gompertz model (Supporting Information).

We chose weakly informative priors to incorporate our understanding of
plausible population dynamics 
(\citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Information). 
For $\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at
values above two (Fig.~\ref{fig:priors}) --- a slightly less informative prior
than suggested by \citet{fernandez1998}. This prior gives only a $7.7$\%
probability that $\nu < 10$ but constrains the sampling sufficiently to avoid
wandering off towards infinity --- above approximately $\nu = 20$ the
t distribution is so similar to the normal distribution that time series of the
length considered here are unlikely to be informative about the precise value
of $\nu$ (Fig.~\ref{fig:didactic}).

We fit alternative population models to test if four key phenomena
systematically changed our conclusions. (1) Autocorrelation has been suggested
as a reason for increased observed variability of abundance time series through
time, which could create apparent heavy tails \citep{inchausti2002}; therefore,
we fit a model that modelled the serial correlation in the residuals. (2)
Previous work has modelled abundance or growth rates without accounting for
density dependence \citep{halley2002,segura2013}; therefore, we fit a simpler
model in which we assumed density independence. (3) Observation error could
bias parameter estimates \citep{knape2012} or mask our ability to detect heavy
tails \citep{ward2007}; therefore, we fit a model where we allowed for a fixed
quantity of observation error ($0.2$ standard deviations on a log scale). (4)
The Gompertz model assumes that population growth rate declines linearly with
log abundance. We also fit an alternative model, the Ricker-logistic model,
which assumes that population growth rate declines linearly with abundance
itself (Supporting Information).

@@

We investigated possible covariates of heavy-tailed population dynamics
visually and through multilevel modelling. We plotted characteristics of the
time series ($\sigma$, $\lambda$, $b$, and time-series length) along with two
life-history characteristics (body length and maximum lifespan obtained from
\citet{brook2006a}) against our estimated probability of heavy tails, Pr$(\nu
< 10) > 0.5$. We formally investigated these relationships by fitting beta
regression multilevel models. For covariates that were derived from Gompertz
model parameter estimates, we incorporated standard deviations around the
means. To account for broad patterns of phylogenetic relatedness, we allowed
for varying intercepts at the taxonomic class, order, and species level
(Supporting Information).

@@

Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy
directions. First, ecological resource management can draw from other
disciplines that focus on heavy tails. For example, earthquake preparedness and
response is focussed on black swan events. Similarly to ecological black swans,
we can rarely predict the specific location and timing of large earthquakes.
But, earthquake preparedness involves spatial planning based on forecast
probabilities to focus early detection efforts and develop disaster response
plans (e.g.~REFERENCE?). A similar focus might benefit resource management once
our ability to predict the spatial probability and covariates of ecological
black swans improves. Second, we can develop management policy that is robust
to heavy tails. For example, setting target population abundances that are
appropriately set back from critical limits may buffer black swan events
\citep[e.g.][]{caddy1996}, and maintaining genetic, phenotypic, and behavioural
diversity may allow some components of populations persist when others are
affected by disease or extreme environmental forces
\citep[e.g.][]{schindler2010, anderson2014}. Finally, extreme and unexpected,
surprising, or counterintuitive ecological dynamics offer a tremendous
opportunity to learn about ecological systems, evaluate when our models break
down, and adjust future management policy \citep{doak2008, pine-iii2009,
  lindenmayer2010}. In recent decades ecology has moved toward focussing on
aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a,
  thompson2013}. Our results suggest that an added focus on ecological extremes
represents the next frontier, particularly in the face of increased climate
extremes \citep{meehl2004,ipcc2012}.

@@

\section{Supporting Information}

The following supporting information is available online for this article:\\
Figures S1--S11\\
Table S1\\
Example Stan code for a heavy-tailed Gompertz model and multilevel beta
regression\\
The GPDD IDs used in our analysis\\
All code and data to recreate our analysis are available at:\\
\url{https://github.com/seananderson/heavy-tails}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{../analysis/correlates-p10.pdf}
\caption{
Potential covariates of heavy-tailed population dynamics (indicated by a high
probability that $\nu < 10$). Shown are (a--c) parameters from the Gompertz
heavy-tailed population model ($\sigma$, $\lambda$, $b$), (b) number of time
steps, (c) body length, and (d) lifespan. For the Gompertz parameters, $\sigma$
refers to the scale parameter of the Student-t process-noise distribution,
$\lambda$ refers to the expected log abundance at the next time step at an
abundance of one, $b$ refers to the density dependence parameter ($1$ is
maximally density independent, $0$ is maximally density dependent, and $<0$ is
inversely density dependent).
%For parameter values, points and lines represent medians and 50\% credible
%intervals of the posterior.
Circles representing a few sharks, crustaceans, and gastropods are filled in
white. Median and 90\% credible interval posterior predictions of a beta
regression multilevel model are shown in panels a and d where there was a high
probability the slope coefficient was different from zero (Supporting
Information).
}
\label{fig:correlates}
\end{center}
\end{figure}
